A meta analysis of tournaments and an evaluation of performance in the Iterated Prisoner's Dilemma.

### 2019-12-28 paper/main.tex
factors/measures to feautures in the text

The aim of this section is to explore the factors that contribute to a
strategy's successful performance. The factors explored are measures regarding a
strategy's behaviour, along with measures regarding the tournaments the
strategies competed in. These are given in Table~\ref{table:manual_measures}.

@@

APL makes use of classifiers to classify strategies according to
various dimensions. These determine whether
a strategy is stochastic or deterministic, whether it makes use of the number of
turns or the game's payoffs. The memory usage measure is calculated as the
memory size of strategy (which is specified in the strategies implementation
in the APL) divide by the number of turns. For example, Evolved
FSM 16 Noise 05 has a memory size of 16 and participated in a tournament where
$n$ was 134. In the given tournament Evolved FSM 16 Noise 05 has a memory usage
of 0.119. For tournaments with a probabilistic ending the number of turns was
not collected, so the memory usage measure is not used for probabilistic ending
tournaments. The SSE is a measure introduced in~\cite{Knight2019} which shows
how close a strategy is to behaving as a ZDs, and subsequently, in an
extortionate way. The method identifies the ZDs closest to a given strategy and
calculates the algebraic distance between them, defined as SSE. A SSE value of 1
indicates no extortionate behaviour at all whereas a value of 0 indicates that a
strategy is behaving a ZDs. The rest of the factors considered are the $CC$ to
$C$, $CD$ to $C$, $DC$ to $C$, and $DD$ to $C$ rates as well as cooperating
ratio of a strategy. The minimum, maximum, medium and median cooperating ratios
of each tournament are also included, and finally the number of turns, the
number of strategies, the number of repetitions and the probabilities of noise
and the game ending are also included.

Table~\ref{table:correlations} shows the correlation coefficients between the
measures of Table~\ref{table:manual_measures} the median score and the median
normalised rank. Note that the correlation for the classifiers is
not included because they are binary variables and they will be evaluated using a
different method. The correlation coefficients for all the measures in
Table~\ref{table:manual_measures} against themselves have also been calculated and a
graphical representation can be found in the Appendix~\ref{app:correlations}.

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[!htbp]
    \begin{center}
    \resizebox{.9\textwidth}{!}{
        \input{correlations_table.tex}
    }
\end{center}
\caption{Correlations table between the measures of Table~\ref{table:manual_measures}
the normalised rank and the median score.}\label{table:correlations}
\end{table}

In standard tournaments the measures  $CC$ to $C$, $C_r$, $C_r / C_{\text{max}}$
and the cooperating ratio compared to $C_{\text{median}}$ and $C_{\text{mean}}$
have a moderate negative effect on the normalised rank, and a moderate positive
on the median score. The SSE error and the $DD$ to $C$ have the opposite
effects. Thus, in standard tournaments behaving cooperatively corresponds to a
more successful performance. Even though being nice pays off,
that's not true against defective strategies. Cooperating after a mutual
defection lowers a strategy's success.
Figure~\ref{fig:rates_of_winners_in_standard_tournaments} confirms that the
winners of standard tournaments always cooperate after a mutual cooperation and
almost always defects after a mutual defection.

@@

Compared to standard tournaments, in both noisy and in probabilistic ending
tournaments the higher the rates of cooperation the lower a strategy's success
and median score. A strategy would want to cooperate less than both
the mean and median cooperator in such settings. In probabilistic ending
tournaments the correlations coefficients have a larger values, indicating a
stronger effect. Thus a strategy will be punished more by it's cooperative
behaviour in probabilistic ending environments, this was seen in Section~\ref{section:evaluation_of_performance}
as well. The distributions of the $C_r$ of the winners in
both tournaments is given by Figure~\ref{fig:c_r_distributions}. It confirms
that the winners in noisy tournaments cooperated less than 35\% of the times
and in probabilistic ending tournaments less than 10\%.
In noisy probabilistic ending tournaments and in over all the tournaments' results,
the only measures that had a moderate affect are $C_r/C_{\text{mean}},
C_r/C_{\text{max}}$ and $C_r$. In such environments cooperative behaviour
appears to be punished by not as much as in noisy and probabilistic ending
tournaments.

@@

The importance that the measures of Table~\ref{table:manual_measures} had on
each classification task; to which cluster a performance was assigned to based
on the normalised rank, and their normalised rank and median score have been
calculated and are given by Figures~\ref{fig:clustering_importance_standard},
\ref{fig:clustering_importance_noise},~\ref{fig:clustering_importance_probend},
\ref{fig:clustering_importance_probend_noise}
and~\ref{fig:clustering_importance_overall}. These show that the classifiers
stochastic, make use of game and make use of length have no significant effect,
and several of the measures that are highted by the importance are inline with
the correlation results. Moreover, the smoothing parameter \(k\) appears to no
have a significant effect either. The most important measures based on the
random forest analysis were $C_{r} / C_{median}$ and $C_r / C_{mean}$.

@@

The effect of both these measures can be further explored. In
Figure~\ref{fig:mean_median_std} the distributions of \(C_r / C_{\text{mean}}\)
and \(C_r / C_{\text{median}}\) are given for the winners in standard tournaments. A value of \(C_r /
C_{\text{mean}} = 1\) imply that the cooperating ratio of the winner was the
same as the mean/median cooperating ratio of the tournament. In standard tournaments, the mean
for both ratios is 1. Therefore, an effective strategy in standard tournaments
was the mean/median cooperator of its respective tournament. In comparison,
Figure~\ref{fig:mean_median_noisy} shows the distributions of the measures for
the winners in noisy tournaments where the mean is at 0.67. Thereupon the winners
cooperated 67\% of the times the mean/median cooperator did. This analysis is
applied to the rest of the tournaments and the distributions are given by
Figures \ref{fig:mean_median_probend}, \ref{fig:mean_median_probend_noisy} and
\ref{fig:mean_median_overall}. In a tournament with noisy and a probabilistic
ending the winners cooperated 60\%, whereas in settings that the type of the
tournament can vary between the types considered in this work the winners
cooperated 67\% of the times the mean or median cooperator did. Finally, in
probabilistic ending tournament it has already been mentioned that defecting
strategies prevail and this result is once again confirmed in this section.

@@

In this section the effect of several measures, regarding a strategy's behaviour
and the tournament in which it participated on its performance were presented.
This was done using two approaches. Correlation coefficients and a random forest
analysis. The results of these are summarised in the following section.

@@

Section~\ref{section:evaluation_of_performance}, covered an analysis of
performance based on several measures associated with a strategy and with the
environments it was competing. The results of this analysis showed that a
strategy's characteristics such as whether or not it's stochastic, and the information it
used regarding the game had no effect on the strategy's success. The most
important factors have been those that compared the strategy's behaviour to it's
environment. The cooperating ratio of the strategy compared to the mean and
median cooperator was highlighted as the most important feature in the analysis.
More specifically, if a strategy were to enter a tournament with a theory of
mind of its environment it would choose to be the median cooperator in standard
tournaments, to cooperate 10\% in probabilistic ending tournaments and 
60\% in noisy and noisy probabilistic
tournaments of the times the median cooperator did. Lastly, if a strategy was aware of the opponents but not of the
setting on the tournament, a strategy would be more likely to be successful if
it were to identify the median cooperator and cooperated 67\% of the times that
they did.

@@

\begin{table}[h]
    \begin{center}
    \resizebox{.7\textwidth}{!}{
    \begin{tabular}{llc}
    \toprule
measure & measure explanation \\
 \midrule
stochastic                  & If a strategy is stochastic \\
makes use of game           & If a strategy makes used of the game information  \\
makes use of length         & If a strategy makes used of the number of turns \\
memory usage                & The memory size of a strategy divided by the number of turns \\
SSE                         & A measure of how far a strategy is from extortionate behaviour \\
$C_{\text{max}}$          & The biggest cooperating rate in the tournament \\
$C_{\text{min}}$          & The smallest cooperating rate in the tournament \\
$C_{\text{median}}$       & The median cooperating rate in the tournament \\
$C_{\text{mean}}$         & The mean cooperating rate in the tournament \\
$C_r$ / $C_{\text{max}}$    & A strategy's cooperating rate divided by the maximum \\
$C_r$ / $C_{\text{min}}$    & A strategy's cooperating rate divided by the minimum \\
$C_r$ / $C_{\text{median}}$ & A strategy's cooperating rate divided by the median \\
$C_r$ / $C_{\text{mean}}$   & A strategy's cooperating rate divided by the mean \\
$C_r$                       & The cooperating ratio of a strategy \\
$CC$ to $C$ rate            & The probability a strategy will cooperate after a mutual cooperation \\
$CD$ to $C$ rate            & The probability a strategy will cooperate after being betrayed by the opponent \\
$DC$ to $C$ rate            & The probability a strategy will cooperate after betraying the opponent\\
$DD$ to $C$ rate            & The probability a strategy will cooperate after a mutual defection\\
$p_n$                       & The probability of a player's action being flip at each interaction \\
$n$                         & The number of turns \\
$p_e$                       & The probability of a match ending in the next turn \\
$N$                         & The number of strategies in the tournament \\
$k$                         & The number that a given tournament is repeated \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{The measures which are included in the performance evaluation analysis.}
\end{table}

@@

A graphical representation of the correlation coefficients for the measures in
Table~\ref{table:manual_measures}.

\begin{figure}[!htbp]
        \begin{center}
            \includegraphics[width=.75\linewidth]{../images/standard_correlation_plot.pdf}
        \end{center}
        \caption{Correlation coefficients of measures in Table~\ref{table:manual_measures}
        for standard tournaments}
\end{figure}
\begin{figure}[!htbp]
    \begin{center}
        \includegraphics[width=.75\linewidth]{../images/noise_correlation_plot.pdf}
    \end{center}
    \caption{Correlation coefficients of measures in Table~\ref{table:manual_measures}
    for noisy tournaments}
\end{figure}
\begin{figure}[!htbp]
    \begin{center}
        \includegraphics[width=.75\linewidth]{../images/probend_correlation_plot.pdf}
    \end{center}
    \caption{Correlation coefficients of measures in Table~\ref{table:manual_measures}
    for probabilistic ending tournaments}
\end{figure}
\begin{figure}[!htbp]
    \begin{center}
        \includegraphics[width=.75\linewidth]{../images/probend_noise_correlation_plot.pdf}
    \end{center}
    \caption{Correlation coefficients of measures in Table~\ref{table:manual_measures}
    for noisy probabilistic ending tournaments}
\end{figure}
\begin{figure}[!htbp]
    \begin{center}
        \includegraphics[width=.75\linewidth]{../images/merged_correlation_plot.pdf}
    \end{center}
    \caption{Correlation coefficients of measures in Table~\ref{table:manual_measures}
    for data set}
\end{figure}

>>>>

The aim of this section is to explore the features that contribute to a
strategy's successful performance. The features explored are measures regarding a
strategy's behaviour, along with measures regarding the tournaments the
strategies competed in. These are given in Table~\ref{table:manual_measures}.

@@

APL makes use of classifiers to classify strategies according to
various dimensions. These determine whether
a strategy is stochastic or deterministic, whether it makes use of the number of
turns or the game's payoffs. The memory usage measure is calculated as the
memory size of strategy (which is specified in the strategies implementation
in the APL) divide by the number of turns. For example, Evolved
FSM 16 Noise 05 has a memory size of 16 and participated in a tournament where
$n$ was 134. In the given tournament Evolved FSM 16 Noise 05 has a memory usage
of 0.119. For tournaments with a probabilistic ending the number of turns was
not collected, so the memory usage measure is not used for probabilistic ending
tournaments. The SSE is a measure introduced in~\cite{Knight2019} which shows
how close a strategy is to behaving as a ZDs, and subsequently, in an
extortionate way. The method identifies the ZDs closest to a given strategy and
calculates the algebraic distance between them, defined as SSE. A SSE value of 1
indicates no extortionate behaviour at all whereas a value of 0 indicates that a
strategy is behaving a ZDs. The rest of the features considered are the $CC$ to
$C$, $CD$ to $C$, $DC$ to $C$, and $DD$ to $C$ rates as well as cooperating
ratio of a strategy. The minimum, maximum, medium and median cooperating ratios
of each tournament are also included, and finally the number of turns, the
number of strategies, the number of repetitions and the probabilities of noise
and the game ending are also included.

Table~\ref{table:correlations} shows the correlation coefficients between the
features of Table~\ref{table:manual_measures} the median score and the median
normalised rank. Note that the correlation for the classifiers is
not included because they are binary variables and they will be evaluated using a
different method. The correlation coefficients for all the features in
Table~\ref{table:manual_measures} against themselves have also been calculated and a
graphical representation can be found in the Appendix~\ref{app:correlations}.

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[!htbp]
    \begin{center}
    \resizebox{.9\textwidth}{!}{
        \input{correlations_table.tex}
    }
\end{center}
\caption{Correlations table between the features of Table~\ref{table:manual_measures}
the normalised rank and the median score.}\label{table:correlations}
\end{table}

In standard tournaments the features  $CC$ to $C$, $C_r$, $C_r / C_{\text{max}}$
and the cooperating ratio compared to $C_{\text{median}}$ and $C_{\text{mean}}$
have a moderate negative effect on the normalised rank, and a moderate positive
on the median score. The SSE error and the $DD$ to $C$ have the opposite
effects. Thus, in standard tournaments behaving cooperatively corresponds to a
more successful performance. Even though being nice pays off,
that's not true against defective strategies. Cooperating after a mutual
defection lowers a strategy's success.
Figure~\ref{fig:rates_of_winners_in_standard_tournaments} confirms that the
winners of standard tournaments always cooperate after a mutual cooperation and
almost always defects after a mutual defection.

@@

Compared to standard tournaments, in both noisy and in probabilistic ending
tournaments the higher the rates of cooperation the lower a strategy's success
and median score. A strategy would want to cooperate less than both
the mean and median cooperator in such settings. In probabilistic ending
tournaments the correlations coefficients have a larger values, indicating a
stronger effect. Thus a strategy will be punished more by it's cooperative
behaviour in probabilistic ending environments, this was seen in Section~\ref{section:evaluation_of_performance}
as well. The distributions of the $C_r$ of the winners in
both tournaments is given by Figure~\ref{fig:c_r_distributions}. It confirms
that the winners in noisy tournaments cooperated less than 35\% of the times
and in probabilistic ending tournaments less than 10\%.
In noisy probabilistic ending tournaments and in over all the tournaments' results,
the only features that had a moderate affect are $C_r/C_{\text{mean}},
C_r/C_{\text{max}}$ and $C_r$. In such environments cooperative behaviour
appears to be punished by not as much as in noisy and probabilistic ending
tournaments.

@@

The importance that the features of Table~\ref{table:manual_measures} had on
each classification task; to which cluster a performance was assigned to based
on the normalised rank, and their normalised rank and median score have been
calculated and are given by Figures~\ref{fig:clustering_importance_standard},
\ref{fig:clustering_importance_noise},~\ref{fig:clustering_importance_probend},
\ref{fig:clustering_importance_probend_noise}
and~\ref{fig:clustering_importance_overall}. These show that the classifiers
stochastic, make use of game and make use of length have no significant effect,
and several of the features that are highted by the importance are inline with
the correlation results. Moreover, the smoothing parameter \(k\) appears to no
have a significant effect either. The most important features based on the
random forest analysis were $C_{r} / C_{median}$ and $C_r / C_{mean}$.

@@

The effect of both these features can be further explored. In
Figure~\ref{fig:mean_median_std} the distributions of \(C_r / C_{\text{mean}}\)
and \(C_r / C_{\text{median}}\) are given for the winners in standard tournaments. A value of \(C_r /
C_{\text{mean}} = 1\) imply that the cooperating ratio of the winner was the
same as the mean/median cooperating ratio of the tournament. In standard tournaments, the mean
for both ratios is 1. Therefore, an effective strategy in standard tournaments
was the mean/median cooperator of its respective tournament. In comparison,
Figure~\ref{fig:mean_median_noisy} shows the distributions of the features for
the winners in noisy tournaments where the mean is at 0.67. Thereupon the winners
cooperated 67\% of the times the mean/median cooperator did. This analysis is
applied to the rest of the tournaments and the distributions are given by
Figures \ref{fig:mean_median_probend}, \ref{fig:mean_median_probend_noisy} and
\ref{fig:mean_median_overall}. In a tournament with noisy and a probabilistic
ending the winners cooperated 60\%, whereas in settings that the type of the
tournament can vary between the types considered in this work the winners
cooperated 67\% of the times the mean or median cooperator did. Finally, in
probabilistic ending tournament it has already been mentioned that defecting
strategies prevail and this result is once again confirmed in this section.

@@

In this section the effect of several features, regarding a strategy's behaviour
and the tournament in which it participated on its performance were presented.
This was done using two approaches. Correlation coefficients and a random forest
analysis. The results of these are summarised in the following section.

@@

Section~\ref{section:evaluation_of_performance}, covered an analysis of
performance based on several features associated with a strategy and with the
environments it was competing. The results of this analysis showed that a
strategy's characteristics such as whether or not it's stochastic, and the information it
used regarding the game had no effect on the strategy's success. The most
important features have been those that compared the strategy's behaviour to it's
environment. The cooperating ratio of the strategy compared to the mean and
median cooperator was highlighted as the most important feature in the analysis.
More specifically, if a strategy were to enter a tournament with a theory of
mind of its environment it would choose to be the median cooperator in standard
tournaments, to cooperate 10\% in probabilistic ending tournaments and 
60\% in noisy and noisy probabilistic
tournaments of the times the median cooperator did. Lastly, if a strategy was aware of the opponents but not of the
setting on the tournament, a strategy would be more likely to be successful if
it were to identify the median cooperator and cooperated 67\% of the times that
they did.

@@

\begin{table}[h]
    \begin{center}
    \resizebox{.7\textwidth}{!}{
    \begin{tabular}{llc}
    \toprule
measure & measure explanation \\
 \midrule
stochastic                  & If a strategy is stochastic \\
makes use of game           & If a strategy makes used of the game information  \\
makes use of length         & If a strategy makes used of the number of turns \\
memory usage                & The memory size of a strategy divided by the number of turns \\
SSE                         & A measure of how far a strategy is from extortionate behaviour \\
$C_{\text{max}}$          & The biggest cooperating rate in the tournament \\
$C_{\text{min}}$          & The smallest cooperating rate in the tournament \\
$C_{\text{median}}$       & The median cooperating rate in the tournament \\
$C_{\text{mean}}$         & The mean cooperating rate in the tournament \\
$C_r$ / $C_{\text{max}}$    & A strategy's cooperating rate divided by the maximum \\
$C_r$ / $C_{\text{min}}$    & A strategy's cooperating rate divided by the minimum \\
$C_r$ / $C_{\text{median}}$ & A strategy's cooperating rate divided by the median \\
$C_r$ / $C_{\text{mean}}$   & A strategy's cooperating rate divided by the mean \\
$C_r$                       & The cooperating ratio of a strategy \\
$CC$ to $C$ rate            & The probability a strategy will cooperate after a mutual cooperation \\
$CD$ to $C$ rate            & The probability a strategy will cooperate after being betrayed by the opponent \\
$DC$ to $C$ rate            & The probability a strategy will cooperate after betraying the opponent\\
$DD$ to $C$ rate            & The probability a strategy will cooperate after a mutual defection\\
$p_n$                       & The probability of a player's action being flip at each interaction \\
$n$                         & The number of turns \\
$p_e$                       & The probability of a match ending in the next turn \\
$N$                         & The number of strategies in the tournament \\
$k$                         & The number that a given tournament is repeated \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{The features which are included in the performance evaluation analysis.}\label{table:parameters_summary}
\end{table}

@@

A graphical representation of the correlation coefficients for the features in
Table~\ref{table:manual_measures}.

\begin{figure}[!htbp]
        \begin{center}
            \includegraphics[width=.75\linewidth]{../images/standard_correlation_plot.pdf}
        \end{center}
        \caption{Correlation coefficients of features in Table~\ref{table:manual_measures}
        for standard tournaments}
\end{figure}
\begin{figure}[!htbp]
    \begin{center}
        \includegraphics[width=.75\linewidth]{../images/noise_correlation_plot.pdf}
    \end{center}
    \caption{Correlation coefficients of features in Table~\ref{table:manual_measures}
    for noisy tournaments}
\end{figure}
\begin{figure}[!htbp]
    \begin{center}
        \includegraphics[width=.75\linewidth]{../images/probend_correlation_plot.pdf}
    \end{center}
    \caption{Correlation coefficients of features in Table~\ref{table:manual_measures}
    for probabilistic ending tournaments}
\end{figure}
\begin{figure}[!htbp]
    \begin{center}
        \includegraphics[width=.75\linewidth]{../images/probend_noise_correlation_plot.pdf}
    \end{center}
    \caption{Correlation coefficients of features in Table~\ref{table:manual_measures}
    for noisy probabilistic ending tournaments}
\end{figure}
\begin{figure}[!htbp]
    \begin{center}
        \includegraphics[width=.75\linewidth]{../images/merged_correlation_plot.pdf}
    \end{center}
    \caption{Correlation coefficients of features in Table~\ref{table:manual_measures}
    for data set}
\end{figure}
